Value-Suppressing Uncertainty Maps

### 2017-03-30 paper/figures.tex
Jeff's revisions.

\newcommand{\exampleFig}{
\begin{figure}
	\centering
	\includegraphics[width=0.9\columnwidth]{example.pdf}
	\caption{A comparison between a standard bivariate map and a VSUM. Both use the same visual channels to encode value (position along the Viridis color map) and uncertainty (lightness and saturation), and have an equal standard of perceptual discriminability (at least 18 units of distance in CIELAB color space between colors). However, highly uncertain values result in colors that are very close together in the bivariate case, meaning the full bivariate map is only 9 bins under these constraints--- a 4x4 map would result in colors that are perceptually too close together. By contrast, the VSUM intentionally reduces bins when uncertainty is high, eventually aliasing all highly uncertain values to the same color. This decision affords more distinct colors in other regions of the map, and so an increase in overall bins to 15. The resulting map suppresses the value of uncertain data, but increases the discriminability of data with low uncertainty.}
	\label{fig:example}
\end{figure}
}

\newcommand{\performanceFig}{
	\begin{figure}
		\centering
		\includegraphics[width=0.9\columnwidth]{performance.pdf}
		\caption{
			The number of discrete color categories in traditional bivariate color maps versus VSUMs, assuming value is encoded by the Viridis color map~\protect\cite{garnier2015viridis}, and uncertainty is encoded by saturation/value (whiter values are more uncertain). Since these whiter uncertain colors are closer together in CIELAB space, and VSUMs intentionally require fewer color bins in those regions, VSUMs always have more color categories than the corresponding 2D bivariate map. The red lines denote:\\
		    \textbf{1)} Theoretical 50\% JND from CIELAB specification. \\
			\textbf{2)} 50\% JND from controlled lab study of Mahy et al.~\protect\cite{mahy1994evaluation}.\\
			\textbf{3)} 50\% JND from Szafir et al.\protect\cite{szafir2014adapting}.\\
			\textbf{4)} 50\% JND for small visual targets from Stone et al.~\protect\cite{stone2014engineering}.\\
			\textbf{5)} The threshold resulting in the simplest possible (2x2) bivariate map. \\
	    }
		\label{fig:performance}
	\end{figure}
}

>>>>

\newcommand{\exampleFig}{
\begin{figure}
	\centering
	\includegraphics[width=0.9\columnwidth]{example.pdf}
	\caption{A standard bivariate map and a VSUM. Both use the same visual channels to encode value (position along the Viridis color map) and uncertainty (lightness and saturation), and have an equal standard of perceptual discriminability (at least 18 units of distance in CIELAB color space between colors). However, highly uncertain values result in colors that are very close together in the bivariate case, meaning the full bivariate map is only 9 bins under these constraints\,---\,a 4x4 map would result in colors that are perceptually too close together. By contrast, the VSUM intentionally reduces bins when uncertainty is high, eventually aliasing all highly uncertain values to the same color. This decision affords more distinct colors in other regions of the map, and so an increase in overall bins to 15. The resulting map suppresses the value of uncertain data, but increases the discriminability of data with low uncertainty.}
	\label{fig:example}
\end{figure}
}

\newcommand{\performanceFig}{
	\begin{figure}
		\centering
		\includegraphics[width=0.9\columnwidth]{performance.pdf}
		\caption{
			The number of discrete color categories in traditional bivariate color maps versus VSUMs, assuming value is encoded by the Viridis color map~\protect\cite{garnier2015viridis}, and uncertainty is encoded by saturation/value (whiter values are more uncertain). Since these whiter uncertain colors are closer together in CIELAB space, and VSUMs intentionally require fewer color bins in those regions, VSUMs always have more color categories than the corresponding 2D bivariate map. The red lines denote:
		  \textbf{1)} Theoretical 50\% JND from CIELAB specification.
			\textbf{2)} 50\% JND from controlled lab study of Mahy et al.~\protect\cite{mahy1994evaluation}.
			\textbf{3)} 50\% JND from Szafir et al.\protect\cite{szafir2014adapting}.
			\textbf{4)} 50\% JND for small visual targets from Stone et al.~\protect\cite{stone2014engineering}.
			\textbf{5)} The threshold resulting in the simplest possible (2x2) bivariate map.
	    }
		\label{fig:performance}
	\end{figure}
}

### 2017-03-30 paper/template.tex
Jeff's revisions.

Uncertainty is an inescapable component of collecting, presenting, and using data. A common goal in the communication of uncertainty is promoting \emph{uncertainty-aware decisions}--- that is, the audience should be aware of the risks and rewards of certain decisions, modulate their confidence in their conclusions, and perhaps refrain from making a decision at all if there is too much uncertainty.  A way that designers can contribute to this goal is by ensuring that uncertainty information is \emph{well-integrated} with the rest of the data. That is, it should be difficult to discount or ignore the uncertainty in a dataset. Simultaneous presentation of uncertainty and value necessitates the construction of a bivariate map--- a relation, in terms of visual variables, between 2-tuples $(\text{value}, \text{uncertainty})$ and mark properties. The design of bivariate maps is difficult and, due to the interference and interplay between different visual variables, often leads to maps with few discriminable categories.

In this paper, we present \textbf{Value-Suppressing Uncertainty Maps} (VSUMs) for integrating data and uncertainty information in thematic maps. VSUMs intentionally alias together data values with high uncertainty, affording greater discriminability as uncertainty decreases. Rather than a traditional two-dimensional bivariate map, with differing outputs for each combination of value and uncertainty, VSUMs can be thought of as arcs: as uncertainty increases, values are mapped to smaller and smaller sets of outputs, culminating in a singularity where all inputs are mapped to an identical, highly uncertain mark regardless of their data value. Figure \ref{fig:example} shows an example of a VSUM, compared to a more traditional bivariate map. In this work, we describe the motivations behind the use of VSUMs, examples of their utility for decision-making under uncertainty, and assess VSUMs in a crowd-sourced experiment. Our experimental results indicate that VSUMs promote better integration between uncertainty and data: X.

@@

Despite the acknowledged importance of uncertainty in understanding data, explicit representations of uncertainty are often missing from visualizations~\cite{boukhelifa2009uncertainty}. This is partially due to the complexity of uncertainty as a concept. In typologies from Thomson et al~\cite{thomson2005typology} and Buttenfield \& Beard~\cite{buttenfield1994graphical}, the authors note that many, occasionally contradictory, concepts can fall under the category of ``uncertainty,'' including data quality, sampling error, credibility, and provenance.

In their reviews of the state of the art in uncertainty visualization in the field, Greithe et al.~\cite{griethe2006visualization} and Brodlie et al.~\cite{brodlie2012review} present lists of potential techniques for conveying uncertainty in concert with data. Some of these options (interaction, animation, and sonification) are not applicable for static charts; even for dynamic charts, many users do not interact with charts in sufficient detail to recover uncertainty information~\cite{nyt2016}. While we acknowledge the potential utility of these techniques in uncertainty visualization (for instance, the animation used to convey sampling error in Hullman et al.'s \emph{HOPs}~\cite{hullman2015hypothetical}), we limit the scope of our discussion to techniques which center on static charts.

To support our goal of making uncertainty information well-integrated with data, we focus on the techniques of \emph{juxtaposition} (where uncertainty is presented alongside the data), \textbf{superposition} (where additional glyphs are overlaid on top of the data), or \emph{explicit encoding} (where a spare visual channel, such as color, is used to encode the uncertainty simultaneously with the data). Gleicher et al.~\cite{gleicher2011visual} explore these three strategies from the perspective of affording \emph{comparisons}; here, we examine them from the perspective of information fusion.

@@

\emph{Superposition}, or the overlay of glyphs representing uncertainty on top of a data map, is another common strategy. Error bars are a common example of this strategy: a mark indicating measurement error is superimposed atop a mark representing a point value. Ware's~\cite{ware2009quantitative} \emph{textons} are an example of this strategy applied to thematic maps: monochrome glyphs that can encode an ordinal variable, or a binned quantitative variable. Greithe et al.~\cite{griethe2006visualization} and MacEachren~\cite{maceachren1992visualizing,maceachren1998visualizing} present other examples of superimposed uncertainty in thematic maps, including overlaid grids with ``fuzzy'' lines to indicate uncertainty, or overlaid treemaps with coarser and coarser resolution in uncertain regions. Costs to this approach include obscuring the underlying datamap (which can cause uncertainty information to dominate the display~\cite{brodlie2012review}), potential visual interactions (for instance, simultaneous contrast effects) that can impair the legibility of glyphs, and the necessity of performing spatial binning in order to ensure that there are a discrete number of uncertainty glyphs. How this binning is performed can introduce undesirable artifacts on the resulting visualization, and suppress important signals, or highlight spurious ones \cite{battersby2016shapes}.

@@

A further hurdle is that not all visual variables are well-suited for conveying uncertainty. MacEachren et al.~\cite{maceachren2012visual} evaluate a number of visual variables with respect to their \emph{semiotic fit} for representing uncertainty. They observe that certain visual variables such as blurriness and transparency seem to have a more intuitive connection to uncertainty than other variables such as shape or hue. Unfortunately, Boukhelifa et al.~\cite{boukhelifa2012evaluating} discover that many visual variables habitually used for conveying uncertainty, such as blur and value, are also difficult to estimate. This results in a preference/performance gap where designers must choose between encoding uncertainty in a way that is intuitive but error prone, or use higher fidelity channels that are difficult to interpret.

@@

In for visualizations like choropleth maps, heatmaps, or treemaps, visual channels such as position and length are reserved for data variables other than value (such as geographic location or relative size). In these situations, data value is habitually encoded using color. Colors in such univariate quantitative color maps should be sufficiently far apart as to be perceptually distinguishable~\cite{ware1988color}, and vary in lightness as well as hue in order to afford an implicit ordering of value~\cite{borland2007rainbow,rogowitz2001blair}. Different choices of color maps can highlight different features of the data, and should be chosen with care based on the data types and important features to be visualized~\cite{rogowitz1996not}. Given these considerations, and the perceptual non-orthogonality of color channels, the construction of \emph{bivariate} color maps is difficult. Viewers need to be able to independently assess the variables of interest, but the resulting bivariate map should also be expressive and easy to interpret. Bivariate maps with these constraints are often limited to relatively few categories (say, a 3x3 matrix as in Figure \ref{fig:example}) \cite{robertson1986generation,trumbo1981theory}. In general, the quality of bivariate color maps is a multivariate measure involving consideration of not just the component color channels, but also the interpolation scheme and the color of the surround~\cite{bernard2015survey}. However, even simple bivariate maps can be difficult for the general audience to interpret \cite{wainer1980empirical}.

Dunn~\cite{dunn1989dynamic} attempts to circumvent these issues by dynamically allocating bins to the bivariate map. Since the amount of distinct, separable colors is limited, the dynamic approach allocates bins to regions of the highest utility. For instance, placing the dividing line between bins in areas aligned with a trend line allows the color to encode information about the sign of the residual with respect to each variable. This insight, that the bins in a bivariate map are a limited resource and so should be ``spent'' wisely, drives the creation of VSUMs, which allocate more bins in a bivariate map to regions with less uncertainty.

@@

We present Value-Suppressing Uncertainty Maps, a technique for creating bivariate maps of data \emph{value} and \emph{uncertainty}. These maps circumvent the issues in the creation of traditional bivariate maps by making the following assumptions:

@@

%MC this paragraph is a little iffy, and sort of "the lady doth protest too much" territory. we might be able to get away with "people bin colors all the time. and in fact, you really should be binning your colors anyway."
The first assumption seems to contradict the \emph{expressiveness principle}~\cite{mackinlay1986automating}, where the visual encoding of the data should reflect the composition of the data. For instance, if my data are a collection of measurements of some continuous variable, then it would make sense to encode these data using a continuous \emph{visual} variable, which would then mean an arbitrary number of potential marks. For instance, a bar can have an arbitrary, non-integer heights. However, for many visualizations, the perceptual ambiguity caused by having marks that cannot be reliably disambiguated is higher than the gain from having a visual encoding that maps readily to the backing data. Padilla et al.~\cite{padilla2017evaluating} present one such example, where binned heatmaps results in higher accuracy in tasks than in continuous heatmaps.

In some cases, this second assumption is violated. For instance, an analyst might be interested in ``long tail risks'' or other ``black swan events,'' where the impact of an value, no matter how uncertain, must be considered and planned for~\cite{taleb2011black}. Other analysis tasks (such as filtering out outliers), require increased, rather than decreased, discriminability when uncertainty low. However, for many information fusion tasks, the assumption is that uncertainty is related to data quality, or the uncertainty of particular data values \cite{riveiro2007evaluation}. As with null hypothesis significance testing, the aim would then be to avoid Type I errors, where patterns arising from statistical noise or measurement error are interpreted as significant signals.

@@

@@

\begin{enumerate}
	\item \textbf{Select a visual variable to encode value}. In most cases in this paper, this variable is position along some established color scale.
	\item \textbf{Select a visual variable to encode uncertainty}. We choose variables that have a somewhat high semiotic connection with uncertainty (such as lightness and size), but still have a reasonable number of perceptually discriminable layers (as opposed to other, more intuitive channels such as blur or sketchiness).
	\item \textbf{Select a discriminability threshold}. This is some minimum distance, in perceptual space, between marks. For many visual variables (such as length or hue), the perceptual discriminability between marks has been assessed in prior literature, and so the only step is to choose some value in terms of the just noticeable distance (JND). For less common visual variables, this perceptual space must be built up empirically (as in Demiralp et al.~\cite{demiralp2014learning}).
	\item \textbf{Choose how quickly categories \emph{degrade}}. That is, if there are $n$ bins at some uncertainty level $m$, how many bins should there be when uncertainty increases to level $m+1$? In Figure \ref{fig:example}, this function is $\left \lfloor {\frac{n}{2}}\right \rfloor$, meaning that the number of value bins decreases by half for each level of uncertainty bins. If there is too much degradation, then there are fewer levels of uncertainty. If there is no degradation, the resulting map is just a traditional bivariate map. 
	\item \textbf{Choose the largest number of initial categories $n$ such that the discriminability threshold is met.} For each pair of marks in the map, perform a pairwise comparison of discriminability. If no marks are perceptually ambiguous by the standards of your threshold, then the map is complete. Otherwise, decrement $n$. 
\end{enumerate}

Figure X shows this process in detail, and how it can be used to generate a bivariate map for presenting uncertainty. Javascript code for generating VSUMs given arbitrary d3 quantitative scale functions is available at \url{URL REMOVED FOR REVIEW}.

Correll et al. use a version of VSUMs in their LayerCake genomics visualization tool \cite{correll2015layercake,correll2011visualizing}, where marks representing genomic data with increasing uncertainty are mapped to a smaller and smaller set of increasingly grey colors, creating the effect of uncertain values retreating into ``fog'' while highly certain values remain prominent. Other bivariate visualizations implicitly alias together uncertain values. For instance, if uncertainty is encoded by transparency, a maximally uncertain glyph may be entirely transparent, and so impossible to distinguish from any other maximally uncertain glyph. Other channels with a semiotic connection to uncertainty, such as saturation, value, blur, or size, also have deleterious effects on the disambiguation of colors and shapes. In both of these cases, this property of aliasing is ad hoc, and places no guarantees on the discriminability of colors. The binning and degradation approach of VSUMs makes this choice to alias values explicit to both the designer and the viewer, and results in a bivariate mapping with known perceptual properties.

@@

There are multiple choices that designers must make before creating a VSUM. In particular, they must choose 1) which visual channels map to value and uncertainty, 2) a threshold of perceptual distinguishability (and, implicitly, a metric for evaluating discriminability), and 3) a degradation function to determine how quickly uncertain values are aliased.

Often, the graphical conventions of a chart reserve variables such as position and area for other uses, and determine the visual channel to be used for value. In thematic maps, for instance, value is habitually encoded using color, while vertical and horizontal position are reserved for location. In this paper, we primarily deal with the case where value is encoded using color. While there are some similar conventions for representing uncertainty, generally designers have a greater degree of freedom in choosing a visual channel for uncertainty~\cite{maceachren1992visualizing}. We recommend channels with both a strong semiotic connection to uncertainty as well as a relatively large number of perceptually distinguishable levels. In this paper, we primarily discuss value/saturation (or, equivalently, transparency against a white background) and size. Correll \& Gleicher~\cite{correll2013error} show that even audiences without statistical backgrounds can interpret uncertainty information encoded in these channels.

Another reason for limiting our discussion of bivariate mappings to color and size channels is that the perceptual distinguishability of these channels has been well-studied by prior work. CIELAB and other color spaces are designed with the intention of perceptual uniformity--- that is, distance in the color space should universally correspond with perceptual distance. Recent work~\cite{stone2014engineering,szafir2014adapting} has investigated the discriminability of colors in such spaces not only under laboratory conditions, but also among the disparate display conditions and color vision proficiency of crowdworkers. Similarly, formative work by Cleveland \& McGill~\cite{cleveland1984graphical} assesses the discriminability of marks of different height, and this too has been extended both to the more variable population of crowdworkers, and to the area channel specifically~\cite{heer2010crowdsourcing,talbot2014four}. We use values from these works as benchmarks for selecting thresholds of discriminability. The ``just noticeable difference'' (JND) of these channels provides a useful minimum; in practice, we would wish to create maps with marks that are far more reliably distinguishable than the 50\% or 75\% discriminability rate of a JND. Figure \ref{fig:performance} shows how these benchmarks can be used to select reasonable thresholds. Other factors, such as nameability and aesthetic preference, can be added in as additional weights impacting the desirability of certain color palettes, as in Gramazio et al.~\cite{gramazio2017colorgorical}.

How quickly the number of bins decreases is another parameter of VSUMs. If this degradation is too large, then we may only have a few discriminable levels of uncertainty. If this degradation is too small, then we can be overwhelmed with colors. For instance, if we only decrement by one bin per level of uncertainty, there would be \emph{1,485} different colors in a VSUM with a tolerance threshold of $1$ unit in CIELAB space. In the examples in this paper,  we halve the number of bins at each increasing level of uncertainty. When the initial bins at the highly certain region is a power of $2$ (as in Fig. \ref{fig:example}), this gives a tree structure to the resulting VSUM. As uncertainty increases, bins in the previous level map to one and only one ``parent'' in the next uncertainty level, resulting in fewer crossings of color boundaries among rows.

@@

@@

We performed a crowd-sourced experiment on Amazon's Mechanical Turk in order to evaluate the effectiveness of VSUMs for integrating uncertainty and value information in thematic maps. This focus on integration meant that we limited our experimental tasks to scenarios where the participants needed to consider both value and uncertainty before making a decision. After a pretest and short tutorial, we gave participants two main tasks:

@@

After completion of the main task, we solicited demographics information, and performed a domain-agnostic general risk aversion assay~\cite{mandrik2005exploring}.

@@

We included both a \textbf{gain} framing (given predictions of where your opponent's ships are, where should you launch your missiles?), and a \textbf{loss} framing (given predictions of where your opponent's missiles are, where should you place your ships?) in order to account for systematic differences in human judgments under uncertainty. Tversky \& Kahneman~\cite{tversky1985framing} illustrate that framings in terms of gains or losses produce reliably different outcomes. In general, the function of the value of gains is concave, but is convex for losses~\cite{kahneman1979prospect}. That is, avoiding a large gain is often more valuable than striving for a similarly large gain.

>>>>

Uncertainty is an inescapable component of collecting, presenting, and using data. A common goal in the communication of uncertainty is promoting \emph{uncertainty-aware decisions}: the audience should be aware of the risks and rewards of certain decisions, modulate their confidence in their conclusions, and perhaps refrain from making a decision at all if there is too much uncertainty.  A way that designers can contribute to this goal is by ensuring that uncertainty information is \emph{well-integrated} with the rest of the data. That is, it should be difficult to discount or ignore the uncertainty in a dataset.

Simultaneous presentation of uncertainty and value necessitates the construction of a bivariate map\,---\,a relation, in terms of visual variables, between 2-tuples $(\text{value}, \text{uncertainty})$ and mark properties. The design of bivariate maps is difficult and, due to the interference and interplay between different visual variables, often leads to maps with few discriminable categories.

In this paper, we contribute \textbf{Value-Suppressing Uncertainty Maps} (VSUMs) for integrating data and uncertainty information in thematic maps. VSUMs intentionally alias together data values with high uncertainty, affording greater discriminability as uncertainty decreases. Rather than a traditional two-dimensional bivariate map, with differing outputs for each combination of value and uncertainty, VSUMs can be thought of as arcs: as uncertainty increases, values are mapped to smaller and smaller sets of outputs, culminating in a singularity where all inputs are mapped to an identical, highly uncertain mark regardless of their data value. Figure~\ref{fig:example} shows examples of both a VSUM and a more traditional bivariate map.

We describe the motivations behind the use of VSUMs, examples of their utility for decision-making under uncertainty, and assess VSUMs in a crowdsourced experiment. Our experimental results indicate that VSUMs promote better integration between uncertainty and data: X.

@@

Despite the acknowledged importance of uncertainty in understanding data, explicit representations of uncertainty are often missing from visualizations~\cite{boukhelifa2009uncertainty}. This is partially due to the complexity of uncertainty as a concept. In typologies from Thomson et al.~\cite{thomson2005typology} and Buttenfield \& Beard~\cite{buttenfield1994graphical}, the authors note that many, occasionally contradictory, concepts can fall under the category of ``uncertainty,'' including data quality, sampling error, credibility, and provenance.

Reviewing the state of the art in uncertainty visualization in the field, Greithe et al.~\cite{griethe2006visualization} and Brodlie et al.~\cite{brodlie2012review} present lists of potential techniques for conveying uncertainty in concert with data. Some of these options (interaction, animation, and sonification) are not applicable for static charts; even for dynamic charts, many users do not interact with charts in sufficient detail to recover uncertainty information~\cite{nyt2016}. While we acknowledge the potential utility of these techniques in uncertainty visualization (for instance, the animation used to convey sampling error in Hullman et al.'s \emph{HOPs}~\cite{hullman2015hypothetical}), we limit the scope of our discussion to techniques which center on static charts.

To support our goal of making uncertainty information well-integrated with data, we focus on the techniques of \emph{juxtaposition} (where uncertainty is presented alongside the data), \textbf{superposition} (where additional glyphs are overlaid on top of the data), and \emph{explicit encoding} (where an additional visual channel is used to encode the uncertainty). Gleicher et al.~\cite{gleicher2011visual} explore these three strategies from the perspective of affording \emph{comparisons}; here, we examine them from the perspective of uncertainty visualization.

@@

\emph{Superposition}, or the overlay of glyphs representing uncertainty on top of a data map, is another common strategy. Error bars are a common example of this strategy: a mark indicating measurement error is superimposed atop a mark representing a point value. Ware's~\cite{ware2009quantitative} \emph{textons} are an example of this strategy applied to thematic maps: monochrome glyphs that can encode an ordinal variable, or a binned quantitative variable. Greithe et al.~\cite{griethe2006visualization} and MacEachren~\cite{maceachren1992visualizing,maceachren1998visualizing} present other examples of superimposed uncertainty in thematic maps, including overlaid grids with ``fuzzy'' lines to indicate uncertainty, or overlaid treemaps with coarser and coarser resolution in uncertain regions. Costs to this approach include obscuring the underlying data map (which can cause uncertainty information to dominate the display~\cite{brodlie2012review}), potential visual interactions that can impair the legibility of glyphs (for instance, simultaneous contrast effects), and the necessity of performing spatial binning in order to ensure that there are a discrete number of uncertainty glyphs. How this binning is performed can introduce undesirable artifacts on the resulting visualization, suppressing important signals or highlighting spurious ones~\cite{battersby2016shapes}.

@@

A further hurdle is that not all visual variables are well-suited for conveying uncertainty. MacEachren et al.~\cite{maceachren2012visual} evaluate a number of visual variables with respect to their \emph{semiotic fit} for representing uncertainty. They observe that certain visual variables such as blurriness and transparency seem to have a more intuitive connection to uncertainty than other variables such as shape or hue. Unfortunately, Boukhelifa et al.~\cite{boukhelifa2012evaluating} find that many visual variables habitually used for conveying uncertainty, such as blur and value, are also difficult to estimate. This results in a preference/performance gap where designers must choose between encoding uncertainty in a way that is intuitive but error prone, or use higher fidelity channels that may be more difficult to interpret.

@@

For visualizations such as choropleth maps, heatmaps, or treemaps, visual channels such as position and length are reserved for data variables other than value (such as geographic location or relative size). In these situations, data value is often encoded using color. Colors in such univariate quantitative color maps should be sufficiently far apart as to be perceptually distinguishable~\cite{ware1988color}, and vary in lightness as well as hue in order to afford an implicit ordering of value~\cite{borland2007rainbow,rogowitz2001blair}. Different choices of color maps can highlight different features of the data, and should be chosen with care based on the data types and important features to be visualized~\cite{rogowitz1996not}.

Given these considerations, and the perceptual integrality of color channels, the construction of \emph{bivariate} color maps is difficult. Viewers need to be able to independently assess the variables of interest, but the resulting bivariate map should also be expressive and easy to interpret. Bivariate maps with these constraints are often limited to relatively few categories (say, a 3x3 matrix as in Figure \ref{fig:example}) \cite{robertson1986generation,trumbo1981theory}. In general, the quality of bivariate color maps is a multivariate measure involving consideration of not just the component color channels, but also the interpolation scheme and the color of the surround~\cite{bernard2015survey}. However, even simple bivariate maps can be difficult for a general audience to interpret \cite{wainer1980empirical}.

Dunn~\cite{dunn1989dynamic} attempts to circumvent these issues by dynamically allocating bins to the bivariate map. As the number of distinct, separable colors is limited, the dynamic approach allocates bins to regions of the highest utility. For instance, placing the dividing line between bins in areas aligned with a trend line allows the color to encode information about the sign of the residual with respect to each variable. This insight, that the bins in a bivariate map are a limited resource and so should be ``spent'' wisely, drives the creation of VSUMs, which allocate more bins in a bivariate map to regions with less uncertainty.

@@

Value-Suppressing Uncertainty Maps (VSUMs) are a technique for creating bivariate maps of data \emph{value} and \emph{uncertainty}. These maps circumvent the issues in the creation of traditional bivariate maps by making the following assumptions:

@@

%MC this paragraph is a little iffy, and sort of "the lady doth protest too much" territory. we might be able to get away with "people bin colors all the time. and in fact, you really should be binning your colors anyway."
The first assumption seems to contradict Mackinlay's \emph{expressiveness principle}~\cite{mackinlay1986automating}, where the visual encoding of the data should reflect the composition of the data. For instance, if my data are a collection of measurements of some continuous variable, then it would make sense to encode these data using a continuous \emph{visual} variable, which would then mean an arbitrary number of potential marks. For instance, a bar can have an arbitrary, non-integer height. However, for many visualizations, the perceptual ambiguity caused by having marks that cannot be reliably disambiguated is higher than the gain from having a visual encoding that maps readily to the backing data. Padilla et al.~\cite{padilla2017evaluating} present one such example, where binned heatmaps result in higher task accuracy than continuous heatmaps.

In some cases, this second assumption is violated. For instance, an analyst might be interested in ``long tail risks'' or other ``black swan events,'' where the impact of an value, no matter how uncertain, must be considered and planned for~\cite{taleb2011black}. Other analysis tasks (such as filtering out outliers), require increased, rather than decreased, discriminability when uncertainty is high. However, for many information fusion tasks, the assumption is that uncertainty is related to data quality, or the uncertainty of particular data values~\cite{riveiro2007evaluation}. As with null hypothesis significance testing, the aim would then be to avoid Type I errors, where patterns arising from statistical noise or measurement error are interpreted as significant signals.

@@

\subsection{Method}

@@

\begin{enumerate}
	\item \textbf{Select a visual variable to encode value}. In most cases in this paper, this variable is position along some established color scale.
	\item \textbf{Select a visual variable to encode uncertainty}. We prefer variables that have a somewhat high semiotic connection with uncertainty (such as lightness and size), but still have a reasonable number of perceptually discriminable layers (as opposed to other, more intuitive channels such as blur or sketchiness).
	\item \textbf{Select a discriminability threshold}. Choose some minimum distance, in a perceptual space, between marks. For many visual variables (such as length or hue), the perceptual discriminability between marks has been assessed in prior literature, and so the only step is to choose a value in terms of the just noticeable distance (JND). For less common visual variables, a perceptual model may be built empirically (as in Demiralp et al.~\cite{demiralp2014learning}).
	\item \textbf{Choose how quickly categories \emph{degrade}}. That is, if there are $n$ bins at some uncertainty level $m$, how many bins should there be when uncertainty increases to level $m+1$? In Figure \ref{fig:example}, this function is $\left \lfloor {\frac{n}{2}}\right \rfloor$, meaning that the number of value bins decreases by half for each level of uncertainty bins. If there is too much degradation, then there are fewer levels of uncertainty. If there is no degradation, the resulting map is just a traditional bivariate map.
	\item \textbf{Choose the largest number of initial categories $n$ such that the discriminability threshold is met.} For each pair of marks in the map, perform a pairwise comparison of discriminability. If no marks are perceptually ambiguous by the standards of your threshold, the map is complete. Otherwise, decrement $n$.
\end{enumerate}

Figure X shows this process in detail, and how it can be used to generate a bivariate map for presenting uncertainty. JavaScript code for generating VSUMs given arbitrary d3 quantitative scale functions is available at \url{URL REMOVED FOR REVIEW}.

Correll et al.~\cite{correll2015layercake,correll2011visualizing} use a pre-cursor of VSUMs in their LayerCake genomics visualization tool, where marks representing genomic data with increasing uncertainty are mapped to a smaller and smaller set of increasingly grey colors, creating the effect of uncertain values retreating into ``fog'' while highly certain values remain prominent. Other bivariate visualizations implicitly alias together uncertain values. For instance, if uncertainty is encoded by transparency, a maximally uncertain glyph may be entirely transparent, and so impossible to distinguish from any other maximally uncertain glyph. Other channels with a semiotic connection to uncertainty, such as saturation, value, blur, or size, also have deleterious effects on the disambiguation of colors and shapes. In both of these cases, the property of aliasing is \emph{ad hoc}, and places no guarantees on the discriminability of colors. The binning and degradation approach of VSUMs makes the choice to alias values explicit to both the designer and the viewer, and results in a bivariate mapping with known perceptual properties.

@@

There are multiple choices that designers must make before creating a VSUM. In particular, they must choose 1) which visual channels map to value and uncertainty, 2) a threshold of perceptual distinguishability (and, implicitly, a corresponding metric), and 3) a degradation function to determine how quickly uncertain values are aliased.

Often, the graphical conventions of a chart reserve variables such as position and area for other uses, and determine the visual channel to be used for value. In thematic maps, for instance, value is regularly encoded using color, while vertical and horizontal position are reserved for location. In this paper, we primarily deal with the case where value is encoded using color. While there are some similar conventions for representing uncertainty, generally designers have a greater degree of freedom in choosing a visual channel for uncertainty~\cite{maceachren1992visualizing}. We recommend channels with both a strong semiotic connection to uncertainty as well as a relatively large number of perceptually distinguishable levels. Here, we focus on value/saturation (or, equivalently, transparency against a white background) and size. Correll \& Gleicher~\cite{correll2013error} show that even audiences without statistical backgrounds can interpret uncertainty information encoded in these channels.

Another reason for limiting our discussion of bivariate mappings to color and size channels is that the perceptual distinguishability of these channels has been well-studied by prior work. CIELAB and other color spaces are designed to approximate perceptual uniformity\,---\,that is, distance in the color space should universally correspond with perceptual distance. Recent work~\cite{stone2014engineering,szafir2014adapting} has investigated the discriminability of colors in such spaces not only under laboratory conditions, but also among the disparate display conditions and color vision proficiency of crowdworkers. Similarly, formative work by Cleveland \& McGill~\cite{cleveland1984graphical} assesses the discriminability of marks of different height, and this too has been extended both to a larger population of crowdworkers, and to the area channel specifically~\cite{heer2010crowdsourcing,talbot2014four}.

We use values from these works as benchmarks for selecting thresholds of discriminability. The ``just noticeable difference'' (JND) of these channels provides a useful minimum; in practice, we would wish to create maps with marks that are far more reliably distinguishable than the 50\% or 75\% discriminability rate of a JND. Figure~\ref{fig:performance} shows how these benchmarks can be used to select reasonable thresholds. Other factors, such as nameability and aesthetic preference, can be added in as additional weights impacting the desirability of certain color palettes, as in Gramazio et al.~\cite{gramazio2017colorgorical}.

How quickly the number of bins decreases is another parameter of VSUMs. If this degradation is too large, then we may only have a few discriminable levels of uncertainty. If this degradation is too small, then we can be overwhelmed with colors. For instance, if we only decrement by one bin per level of uncertainty, there would be \emph{1,485} different colors in a VSUM with a tolerance threshold of $1$ unit in CIELAB space. In the examples in this paper,  we halve the number of bins at each increasing level of uncertainty. When the initial bins at the highly certain region is a power of $2$ (as in Fig. \ref{fig:example}), this imparts a tree structure to the resulting VSUM. As uncertainty increases, bins in the previous level map to one and only one ``parent'' in the next uncertainty level, resulting in fewer crossings of color boundaries among rows.

@@

TODO.

@@

We performed a crowdsourced experiment on Amazon's Mechanical Turk in order to evaluate the effectiveness of VSUMs for integrating uncertainty and value information in thematic maps. This focus on integration meant that we limited our experimental tasks to scenarios where the participants needed to consider both value and uncertainty before making a decision. After a pretest and short tutorial, we gave participants two main tasks:

@@

After completion of the main task, we solicited demographic information, and performed a domain-agnostic general risk aversion assay~\cite{mandrik2005exploring}.

@@

We included both a \textbf{gain} framing (given predictions of where your opponent's ships are, where should you launch your missiles?), and a \textbf{loss} framing (given predictions of where your opponent's missiles are, where should you place your ships?) in order to account for systematic differences in human judgments under uncertainty. Tversky \& Kahneman~\cite{tversky1985framing} illustrate that framings in terms of gains or losses produce reliably different outcomes. In general, the function of the value of gains is concave, but is convex for losses~\cite{kahneman1979prospect}. That is, avoiding a large loass is often more valuable than striving for a similarly large gain.
